---
title: "GWASpoly Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GWASpoly Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE,collapse=FALSE,comment="##",fig.width=5,fig.height=5)
```

## Overview
The following table outlines the basic workflow:

| Function      | Purpose     |
|:--------------|:------------|
|read.GWASpoly|Read in the marker and phenotype data|
|set.K|Calculate the kinship matrix|
|set.params|Specify additional model parameters|
|GWASpoly|Perform the mixed model association test|
|qq.plot|Check inflation of p-values|
|set.threshold|Establish QTL detection threshold|
|manhattan.plot|Visualize p-value results|
|get.QTL|Extract list of significant markers|
|write.GWASpoly|Write results to file|

##1. Read the data
The potato marker (TableS1.csv) and phenotype (TableS2.csv) datafiles that come with the package were published as supplemental files by Rosyara et al. (2016). The `system.file` command is used to obtain the path to these files, which get installed in the subdirectory "extdata." You will not need to do this when analyzing your own data; just pass the names of the files as strings.
```{r}
TableS1 <- system.file("extdata", "TableS1.csv", package = "GWASpoly")
TableS2 <- system.file("extdata", "TableS2.csv", package = "GWASpoly")

library(GWASpoly)
data <- read.GWASpoly(ploidy=4,pheno.file=TableS2,geno.file=TableS1,format="ACGT",n.traits=13,delim=",")
```
The first three columns of the geno.file contain the marker name, chromosome, and map position (in either cM or bp). Subsequent columns contain the marker data, which can be coded using one of three formats. The sample potato dataset is provided in "ACGT" (i.e., nucleotide) format. The other possible formats are "numeric" (0,1,2,..ploidy) and "AB". Missing marker data are imputed with the population mode (most frequent genotype).

The first column of the pheno.file contains the genotype identifier (i.e., the name of the individual or clone), followed by columns of trait data (specify the number of traits with `n.traits`). After the traits can be columns with potential fixed effects. In the potato example, there are 13 traits and 4 columns with population structure covariates.

##2. Set the kinship matrix and other parameters
```{r}
data2 <- set.K(data)
params <- set.params(fixed=c("Grp1","Grp2","Grp3","Grp4"),fixed.type=rep("numeric",4))
```
The function `set.K` is used to set the covariance matrix for a random polygenic effect. (K stands for kinship, but this is a misnomer.) The default approach is $K=MM'$, where $M$ is the centered genotype matrix. In many situations, the polygenic effect is sufficient to control population structure. However, if there is strong stratification, then including population structure covariates may be desirable to further reduce p-value inflation. The function `set.params` can be used for this purpose (shown above), as well as to include other types of fixed effects (e.g., when data are collected in multiple environments).

##3. Testing markers for significance
The following code tells the software to analyze two traits using additive and simplex dominance models. 
```{r}
data3 <- GWASpoly(data2,models=c("additive","1-dom"),traits=c("tuber_shape","tuber_eye_depth"),params=params)
```
For the dominance model, two different analyses are conducted, corresponding to whether the reference (A) or alternate (B) allele is dominant. For "ACGT" marker data, the choice of reference allele is arbtirary and included when results are written to file. 

With any 


